---
title: "Finvars_Cleaning_Script_v01"
author: "Rita M. Ludwig"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
## Load in
#Packages
library(tidyverse)
library(sjlabelled)
library(stringr)
library(textclean)
#Data
load("~/P1_Baseline_Cleaned_2023.Rdata")
#Dictionary
datadict = read.csv('~/P1_baseline_datadictionary_2023.csv')

## Set path and filename to write out final dataset to
path = "~/Cleaned"
filename = "P1_Baseline_Cleaned_Finvars_2023"
```
INPUT: 1) Cleaned dataset.Rdata 2) Data dictionary.csv
OUTPUT: Two new collections of variables (see below for details); cleaned dataset .csv, .dta, .Rdata files.
BEHAVIOR: Cleans financial income variables. For variables that are of number entry type, passes them through a standardized set of cleaning steps to result in a numeric value. For variables that are part of entry-frequency pairs, uses a function to generate a) a monthly figure and b) an annual figure for each variable category.
```{r variable_vectors, include=FALSE}
## Create vectors of the variables to be cleaned. This will be all variables related to personal finances that are of interest for reports, either on their own or as part of calculating some metric such as relationship to the U.S. Census poverty threshold level.
#Vector of all financial variables to be cleaned in this script. Select only financial INCOME vars, not expenses.
finvars = data %>%
  select(., contains("entry") | contains("frequency") | contains("bracket")) %>%
  select(., contains("finances")) %>%
  select(., -contains("healthcare") & -contains("childcare") & -contains("insurance") & -contains("transport") & -contains("emergency")) %>%
  colnames(.)
#Vector of all frequency variables.
freqvars = finvars %>%
  str_subset(., "frequency") %>%
  str_extract(., pattern = "^([^_]*_[^_]*[^_]*)_")
#Vector of all number entry vars. Need this in order to clean the text that participants provided.
entryvars = finvars %>%
  str_subset(., "entry")
#Create another vector of the variables expecting annual figures, since some people provided monthly and that needs to be converted.
annualvars = entryvars %>%
  str_subset(., "income")
```
```{r entry_cleaning, include=FALSE}
## Clean the text entry income variables. Run through a standardized list of cleaning steps, including converting all of the 'cute' ways participants enter numbers. This cleaning saves the updates in-place, meaning that no new 'cleaned' variables are created in the dataset; the original values are overwritten with the cleaned ones.
#Clean the data to make these variables numerical. Converts 'don't know' answers to NA; replace Os substituted for 0s in hundreds; replace commas written after the dollar figure to periods (e.g. "90,00 to 90.00"); replace Ks following numbers with 000s; convert numbers written in English to digits, strip decimal places from numbers.
data = data %>% 
  mutate(across(all_of(entryvars), ~ tolower(.))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "not sure"), NA_real_))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "unknown"), NA_real_))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "don't know"), NA_real_))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "n\\\\a"), NA_real_))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "skip"), NA_real_))) %>%
  mutate(across(all_of(entryvars), ~ replace(., str_detect(., "none"), 0))) %>%
  mutate(across(all_of(entryvars), ~ str_replace_all(., 'OOO', "000"))) %>%
  mutate(across(all_of(entryvars), ~ str_replace_all(., '(,\\d{2})$', '.00'))) %>%
  mutate(across(all_of(entryvars), ~ str_replace_all(., regex('(?<![[:alpha:]])K\\b', ignore_case = TRUE), "000"))) %>%
  mutate(across(all_of(entryvars), ~ str_replace_all(., regex('thousand', ignore_case = TRUE), "000"))) %>%
  mutate(across(all_of(entryvars), ~ str_remove(., pattern = "\\.{1,}[0-9]{2}"))) %>%
  mutate(across(all_of(entryvars), ~ mgsub(., replace_number(seq_len(100)), seq_len(100))))
#Now, detect instances where variables expecting annual figures were provided monthly ones, and calculate their annual estimates.
monthtoannual = function(x) {
  tempdata = data %>% 
    select(., all_of(c("id", x))) %>%
    filter(., str_detect(!!sym(x), "month")) %>%
    mutate(across(all_of(!!x), ~ str_extract(., pattern = "\\d+,?\\d+"))) %>%
    mutate(across(all_of(!!x), ~ str_replace_all(., '(,\\d{2})$', ''))) %>%
    mutate(across(all_of(!!x), ~ str_remove_all(., "\\D+"))) %>%
    mutate(across(all_of(!!x), ~ as.numeric(.)*12)) %>%
    mutate(across(all_of(!!x), ~ as.character(.)))
  data <<- data %>% rows_update(., tempdata, by = "id")
}
#Run the function
map(annualvars, ~monthtoannual(.x))
#Finally, clean all of the income variables so that the number is extracted.
data = data %>%
  mutate(across(all_of(entryvars), ~ str_remove_all(., "\\D+"))) %>%
  mutate(across(all_of(entryvars), ~ as.numeric(.)))
```
```{r frequency_cleaning, include=FALSE}
## Clean the frequency variables and convert them all into a standard metric. Here, generate two new variables each - one for monthly, and one for annual converted amounts.
#This function takes the vector of frequency variable prefixes and runs them through a mathematical conversion to generate monthly and annual figures.
freqconv = function(z) {
  tempnames = data %>% 
  select(., "id", contains(z)) %>%
  select(., -contains("months") & -contains("spouse"))
  tempfreq = str_subset(colnames(tempnames), "frequency")
  tempentr = str_subset(colnames(tempnames), "entry")
  tempnames = tempnames %>% 
    mutate(., "{z}annualy" := case_when(!!sym(tempfreq) == 5 ~ !!sym(tempentr)*12,
                                      !!sym(tempfreq) == 4 ~ !!sym(tempentr)*24,
                                      !!sym(tempfreq) == 3 ~ !!sym(tempentr)*26,
                                      !!sym(tempfreq) == 2 ~ !!sym(tempentr)*52,
                                      !!sym(tempfreq) == 1 ~ !!sym(tempentr)*1
                                      ))
   tempnames = tempnames %>% 
    mutate(., "{z}monthly" := case_when(!!sym(tempfreq) == 5 ~ !!sym(tempentr)*1,
                                      !!sym(tempfreq) == 4 ~ !!sym(tempentr)*2,
                                      !!sym(tempfreq) == 3 ~ !!sym(tempentr)*2.17,
                                      !!sym(tempfreq) == 2 ~ !!sym(tempentr)*4.35,
                                      !!sym(tempfreq) == 1 ~ !!sym(tempentr)/12
                                      ) %>% ceiling(.))
   tempnames = tempnames %>% select(-c(tempfreq, tempentr))
   data <<- data %>% left_join(., tempnames, by = "id")
}
#Run the function
map(freqvars, ~freqconv(.x))
```
Save cleaned data files.
```{r save the data, include=FALSE}
write_csv(data, file.path(path, paste0(filename,".csv")))
write_stata(data, file.path(path, paste0(filename,".dta")), drop.na = FALSE, version = 14)
save(data, file = file.path(path, paste0(filename,".Rdata")))
```